---
title: "Causal Inference for Conservation Biologists"
author: "Villaseñor-Derbez, J.C."
date: "22 de julio de 2017"
output: 
  html_notebook:
    fig_caption: yes
    toc: yes
    toc_float: yes
    toc_collapse: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
suppressPackageStartupMessages({
  library(piecewiseSEM)
  library(magrittr)
  library(tidyverse)
  library(stargazer)
  library(broom)
})
```

```{r}
storks <- read.csv("Data/storks.csv")
```

# About the course

This course was part of the pre-conference workshops at the International Congress for Conservation Biology, in Cartagena, Colombia. The course was given by Dr. Achaz von Hardenberg [\@achazhardenberg](https://twitter.com/achazhardenberg).

Subjects on causal inference, path analalysis and bayesian inference were covered.

# Storks deliver babies (p = 0.0079)

> The scientist's mantra: *"Correlation does not imply causation"*

## Plot

```{r}
ggplot(data = storks, mapping=aes(x = Storks, y = Birth)) +
  geom_point() +
  geom_smooth(method = "lm", color = "black") +
  theme_bw() +
  labs(x = "Number of breeding stork pairs", y = "Human birth rate (thousands/year)")
```

## Linear model

```{r, results = "asis"}
lm(Birth ~ Storks, data = storks) %>% 
  stargazer(single.row = T, type = "html", dep.var.labels = "Human birth rate *thousands / year)")
```

## Other variables

Other variables appear to be correlated, so we can inspect their correlation and include relevant ones in the model.

```{r}
library(corrplot)

corrplot(cor(storks[,-1]), method="ellipse", type="lower")
```

```{r, results = "asis", message = F}
lm1 <- lm(Birth ~ Storks, storks)
lm2 <- lm(Area ~ Storks, storks)
lm3 <- lm(Birth ~ Area, storks)
lm4 <- lm(Birth ~ Area + Storks, storks)

stargazer(lm1, lm2, lm3, lm4, type ="html", single.row = T, add.lines = list(AIC = c("AIC", formatC(AIC(lm1, lm2, lm3, lm4)[,2], digits = 2, format = "f"))))

```

After including `Area`, for example, we observe that the effect of `Storks` significantly decreases.

> Correlation imples and unresolved causal structure

Once we have identified processes, we can start to imply causation:

> Causality always imples a completely resolved correlation structure

# Path analysis

Limitations of commonly employed statistical methods (*i.e.* multiple linear regression):

- Can only analyze one dependent variable at a time

- A particular variable can either be a predictor or a response

*"Path analysis is an extension of multiple regression, and was developed to overcome these limitations"*

In path analysis, we call each variable a vertex (or vertices, in plural) and their connections are called edges. Directed graphs include directions (*i.e.* the edges are arrow that define direction). In Indirect graphs, edges only connect vertices, but do not specify the direction of influence.

A full explanation on the grammar of path analysis is found in [Achaz's materials]("Causal inference Cartagena .pdf")

From his model:

```{r}
DiagrammeR::grViz("
      digraph boxes_and_circles{

# Define nodes
node [shape = square
      penwidth = 2,
      fontsize = 24]

A; B; C; D; E; F

# Add edge statements
#Advaance classes
A -> B
B -> C
B -> D
E -> C
E -> D
F -> E
}
      ", height = 500)
```

A is:

- direct cause of B

- indirect cause of C and F

B is:

- caused by A

C and D are:

- indirectly independent on A and F

- directly dependent on B and E

E is:

- directly dependent on F

F is:

- direct cause of E

- Indirect cause of C and F

In the case above, B is an active vertex, becauser it is bot an effect (of A) and a cause (of C or D). Similarly, D is an inactive vertez, which is only an effect of 1 or more other vertices, but not a cause of anything. Often, we try to understand relationships between D and A, leaving out B.

### Exercise

Drawa causal graph with 6 variables:

- (fromAtoF)where:

- A is direct cause of B

- C is a causal child of B

- D is directly dependent on B and C

- F is directly dependent on D and E

```{r}
DiagrammeR::grViz("
      digraph boxes_and_circles{

# Define nodes
node [shape = square
      penwidth = 2,
      fontsize = 24]

A; B; C; D; E; F

# Add edge statements
#Advaance classes
A -> B
B -> C
C -> D
B -> D
D -> F
E -> F

}
      ", height = 500)
```

To start transforming this into statistical models, we perform d-separation:

1. First, we identify all pairs of not adjacent vertices (*i.e. not connected by a direct arrow*)

```{r}
d_separ <- data.frame(Pairs = c("C,D","A,C", "A,D", "A,E", "A,F", "B,E","B,F", "C,F","D,F"))

d_separ
```

2. Then, we identify the parents of any of the vertices of each pair

```{r}

d_separ <- data.frame(Pairs = c("C,D","A,C", "A,D", "A,E", "A,F", "B,E","B,F", "C,F","D,F"),
                      Parents = c("B,E", "B", "B,E", "F", "None", "A, F", "A", "B,E", "B,E"))

d_separ
```

3. The D-separation statements are therefore:

```{r}
d_separ %>% 
  mutate(D_separation = paste0("(", Pairs, ")", ", {", Parents, "}"))
```

The number of elements in the basis test should be given by

$$\frac{V!}{2(V-2)!}-A$$

In R code, we can define a function for this as:

```{r}
numbers <- function(V, A){
  n <- (factorial(V)/(2*factorial((V-2))))-A
  return(n)
}
```

Where:

- V is the number of vertices

-A is the number of arrows

4. Test probabilistic conditionals of every d-separation statement

```{r}
d_separ %>% 
  mutate(D_separation = paste0("(", Pairs, ")", ", {", Parents, "}"),
         Claims = c("D ~ B + E + C", "C ~ B + A", "D ~ B + E + A", "E ~ F + A", "F ~ A", "E ~ A + F + B", "F ~ B + A", "F ~ B + E + C", "F ~ B + E + D"))

```

We then use Fisher's C test to test the composite probability of the whole set.
 
# Re-visit storks and births
 
With a model of the sape:
 
 
 
```{r}
DiagrammeR::grViz("
      digraph boxes_and_circles{

# Define nodes
node [shape = square
      penwidth = 2,
      fontsize = 24]

A; S; B; I

# Add edge statements
#Advaance classes
A -> S
A -> B
B -> I

}
      ", height = 500)
```

Where:

-  A = Area

- S = Storks

- B = Birth rate

- I = Inhabitants

Using the formula above, we know that we need to identify `r numbers(4, 3)` pairs.

Using the steps above, we must identify all non-adjacent pairs

```{r}
storke_pairs <- data.frame(Pairs = c("A, I", "S, B", "S,I"))

storke_pairs
```

Identify all parents of each pair

```{r}
storke_pairs <- data.frame(Pairs = c("A, I", "S, B", "S,I"),
                           Parents = c("B", "A", "A, B"))

storke_pairs
```

The d-separation statements are then:
```{r}
storke_pairs %>% 
  mutate(D_separation = paste0("(", Pairs, ")", ", {", Parents, "}"))
```

Test the probabilistic conditional dependence of every d-statement with the following claims:

```{r}
storke_pairs %>% 
  mutate(D_separation = paste0("(", Pairs, ")", ", {", Parents, "}"),
         Claims = c("I ~ B + A", "B ~ A + S", "I ~ A + B + I"))
```

Evaluate the models:

```{r, results = "asis"}

colnames(storks) <- c("Country", "A", "S", "I", "B")
path1 <- lm(I ~ B + A, storks)
path2 <- lm(B ~ A + S, storks)
path3 <- lm(I ~ A + B + S, storks)

stargazer(path1, path2, path3, type = "html", report = c("vcp"), single.row = T)
```

Using Fisher's C:

$$C = -2\sum_{i = 1}^{k{}} log(p_i)$$
Where:

-  p is the rpobability of the term for wich we are testing independence (*i.e.* the other vertix from the non-adjancet verrix pairs)

C follows a $\chi^2$ distribution, with 2k degrees of freedom: `1-pchisq(C,2*k)`

```{r}
extract_p <- function(model, coeff){
  tidy(model) %>%
    filter(term == coeff) %>%
    select(p.value)
}

ppath1 <- extract_p(path1, "A")
ppath2 <- extract_p(path2, "S")
ppath3 <- extract_p(path3, "S")

C <- rbind(ppath1, ppath2, ppath3) %>% 
  mutate(logs = log(p.value)) %$%
  sum(logs)

C <-  -2*C

test <- 1-pchisq(C,2*3)
```


Thus, the probability is *p = * `r test`.

# Some R-packages for all this

## `piecewiseSEM`

Here, we only specify the dependent paths as within an `lm()` call. For example, in this model, I is caused by B, so we call `lm(B~I)`. We include each lm call inside a list, and pass that to `sem.fit`, which will identify the missing paths, and do all the calculations.

```{r}
storks_model <- list(
  lm(S ~ A, storks),
  lm(B ~ A, storks),
  lm(I ~ B, storks)) %>% 
  sem.fit(data = storks, .progressBar = F, conditional = T) %>% 
  knitr::kable()
```

